Response of soil microbiome composition to selection on methane oxidation rate

### 2023-12-22 manuscript/manuscript.Rmd
update ms for submission to SBB

---
title: Response of soil microbiome composition to selection on methane oxidation rate
author: "Andrew H. Morris^1^* and Brendan J. M. Bohannan^1^"
output:
  word_document:
    reference_docx: reference.docx
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

# Materials and Methods

@@

## Statistical Analysis

@@

We estimated the proportion of variation in CH~4~ oxidation rate due to
variation in the microbiome as the regression of divergence between the positive
line and the control on the cumulative selection differential [@falconer1996].
This estimate is analogous
to estimates of "microbiability" from the animal breeding literature, which
quantifies the
variation in a host trait that is due to microbiome variation [@he2022]. 
The slope of the regression of divergence on cumulative selection differential
provides an estimate of realized microbiability (h^2^ &pm; SE).
Divergence was calculated as the mean CH~4~ oxidation rate of the positive
treatment minus the mean CH~4~ oxidation rate of the control in each
passage. The selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage.
Cumulative selection differential was calculated as the sum of the selection
differential from all preceding selection events. We then regressed cumulative
divergence on cumulative selection differential using the 'lm' function. We report
the slope as percent change by back-calculating the percent change from the
log-transformed data into the original units using the formula $(10^\beta - 1) * 100$
where $\beta$ is the slope.

@@

@@

To estimate the proportion of variation in CH~4~ oxidation rate due to
variation in microbiome composition--i.e., microbiability (as described in the Methods; [@he2022]--we regressed divergence between the
positive selection treatment and the control against the cumulative selection
differential. The microbiability was `r myround(tidy(rh_model)[[2, 2]], 2)` &pm;
`r myround(tidy(rh_model)[[2, 3]], 2)`, though this was not significant (F~`r glance(rh_model)$df`,`r glance(rh_model)$df.residual`~ = 
`r myround(glance(rh_model)$statistic, 2)`, 
p = `r myround(glance(rh_model)$p.value, 2)`).

@@

Bray-Curtis dissimilarity of the soil microbiome varied strongly by passage and
weakly by treatment with an interaction between passage and treatment (`r figures("beta", display = "cite")`). 
Passage explained `r myround(100*tidy_permanova[1, 4])`\% of the
variation in Bray-Curtis dissimilarity (F~`r tidy_permanova[1, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[1, 5], 1)`, p = `r myround(tidy_permanova[1, 6], 3)`), treatment explained 
`r myround(100*tidy_permanova[2, 4])`\%
of the variation (F~`r tidy_permanova[2, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[2, 5], 1)`, p = `r myround(tidy_permanova[2, 6], 3)`), and the interaction
between treatment and passage explained 
`r myround(100*tidy_permanova[3, 4])`\% of the variation (F~`r tidy_permanova[3, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[3, 5], 1)`, p = `r myround(tidy_permanova[3, 6], 3)`). There was 
no difference in dispersion between treatments or passages (F~3,44~ = 0.91, p = 0.45). 
Finally, CH~4~ oxidation rate was correlated with Bray-Curtis dissimilarity across both treatments in passage 5 and explained `r myround(dbrda_percent_explained[1], 1)`\% of the variation in Bray-Curtis dissimilarity (dbRDA: F~`r dbrda_aov[1, 1]`,`r dbrda_aov[2, 1]`~ = `r myround(dbrda_aov[1, 3], 2)`, p = `r myround(dbrda_aov[1, 4], 3)`)

@@

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure that they did not have an effect.
To do this, we subset the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the selection treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
selection treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014; @dedysh2016]. Based on this analysis,
it appears that no methanotrophs were enriched in the positive selection treatment.

@@

We used artificial ecosystem selection to estimate the contribution of variation
in microbiome composition to variation in the rate of an ecosystem function,
CH~4~ oxidation in soil, independent of environmental variation. Understanding
how and to what degree microbiome variation contributes to variation in
ecosystem function is important for many reasons. For example, successful
microbiome manipulations require that the manipulated microbiome contribute to
variation in ecosystem function independent of other drivers of ecosystem
variation (such as variation in environmental conditions). This is because the
drivers of variation in ecosystem functions can interact in complicated ways
(`r figures("concept", display = "cite")`);
for example, environmental variation can indirectly contribute by providing
conditions that select for microbial groups that in turn alter the rates of
ecosystem functions. Determining the independent contribution of microbiome
variation to ecosystem function is crucial because if microbiome composition is
driven primarily by environmental conditions, then introducing a desirable taxon
through microbiome manipulation without altering the environment will likely be
unsuccessful at shifting the targeted ecosystem function. 
The artificial selection approach is different from the comparative and
manipulative approaches used in past attempts at answering this question,
because it can control for both the direct effect of environment on function as
well as the indirect effect on environment via its impact on microbiome
assembly.

In our study, we observed an increase in CH~4~ oxidation rate in the
selection treatment relative to the control, which demonstrates that there was a
response to selection. Given that we observed a response to selection, we
conclude that variation in the microbiome contributes to variation in the
CH~4~ oxidation rate independent of the environment. This suggests that
microbiome manipulations could be an effective approach for altering the rate of
CH~4~ oxidation in this soil, and that the artificial selection approach may
be useful in determining the potential for microbiome manipulations for other
functions in other ecosystems.

Given that variation in the microbiome is associated with variation in the rate
of an ecosystem function in our system, a reasonable follow-up question is "how
much variation in ecosystem function is associated with microbiome variation in
this system?" One way to estimate this is to determine how much the recipient
jars resemble the selected donor jars that were used to inoculate them
[@falconer1996]. We can
calculate the response to selection as the difference between two successive
passages in their mean CH~4~ oxidation rate. We will denote this as $R$. We
can also calculate the strength of selection as the difference in mean CH~4~
oxidation rate between the twelve jars in one generation and the three jars chosen
for selection in that generation, which we will call the selection differential
and denote as $S$. If we plot the cumulative $R$ against the cumulative $S$, the
slope of this relationship will equal the proportion of variation explained by
the microbiome. If the change in mean function from passage one to passage two
($R$) is equal to the difference in mean function between the twelve jars in
passage one and the three jars selected to inoculate passage two ($S$), then we
would conclude that 100% of the variation is due to variation in the microbiome.
Likewise, if recipients do not resemble the donors in their mean CH~4~
oxidation rate and simply wander randomly, then we would conclude that all of
the variation is due to the environment or technical variation.

The relationship between microbiome variation and ecosystem function variation
is analogous to the concept of "heritability" [@lynch1998] used by quantitative
geneticists, or more precisely the concept of "microbiability" [@he2022] proposed
by microbiome scientists who study host-associated microbiomes. Although rarely
used in the study of environmental microbiomes, this concept could be very
useful for understanding and manipulating microbially-mediated functions in a
variety of ecosystems. In our experiment, variation in microbiome taxonomic
composition statistically explained (i.e., was associated with) 
`r myround(tidy(rh_model)[[2, 2]]*100, 1)`\% of the
variation we observed in the rate of CH~4~ oxidation, though this was not
significant. However, we did observe a significant divergence between the 
positive selection and control lines, which suggests that the imposed
selection and passaging of microbiomes was sufficient to generate
variation in soil CH~4~ oxidation rate. Future studies with
greater replication could more precisely estimate the microbiability.
This suggests that
there is substantial potential for altering this ecosystem function through
microbiome manipulation in this soil. It is very likely that the
“environmental microbiability” will be different for other ecosystem functions
in this soil and for CH~4~ oxidation in other soils. However, our experiment
demonstrates that this relationship is measurable and provides an example of how
this can be accomplished.

@@

There is increasing interest in using artificial selection for understanding and
manipulating the microbiomes associated with plants and animals (a.k.a., “microbiome
breeding”; @mueller2022). Our study demonstrates that artificial ecosystem
selection can also be an important tool for exploring the relationship between
microbiome composition and ecosystem function in non-host systems. This
approach can provide unique information about the independent contribution of
microbiomes to ecosystem functions. Such information is crucial if we are to
successfully manipulate environmental microbiomes to alter ecosystem functions,
whether to improve crop productivity [@chaparro2012]  or
ameliorate the impacts of environmental change [@jansson2020].

# Acknowledgments

@@

# Competing Interests

@@

# Data Availability Statement

@@

# References

@@

# Figure Captions

>>>>

---
title: Response of soil microbiome composition to selection on methane oxidation rate
author: "Andrew H. Morris^1^* and Brendan J. M. Bohannan^1^"
output:
  word_document:
    reference_docx: reference.docx
    number_sections: true
bibliography: bibliography.bib
csl: soil-biology-and-biochemistry.csl
---

@@

# Materials and methods

@@

## Statistical analysis

### Response to selection

@@

### Microbiability

We estimated the proportion of variation in CH~4~ oxidation rate due to
variation in the microbiome by comparing the response to selection to the 
imposed selection differential [@falconer1996].
This estimate is analogous
to estimates of "microbiability" from the animal breeding literature, which
quantifies the
variation in a host trait that is due to microbiome variation [@he2022].

For this purpose, the response to selection was calculated as the divergence between the selection treatment and the control. Divergence was the difference between the mean CH~4~ 
oxidation rate of the positive
treatment and the mean CH~4~ oxidation rate of the control in each
passage. In addition, the selection differential was calculated as the difference between
the mean of the three selected jars and the mean of all twelve jars in a passage. The slope of the divergence over the cumulative selection differential
provides an estimate of realized microbiability (h^2^ &pm; SE).
We regressed cumulative
divergence on cumulative selection differential using the 'lm' function. We report
the slope as percent change by back-calculating the percent change from the
log-transformed data into the original units using the formula $(10^\beta - 1) * 100$
where $\beta$ is the slope.

### Alpha and beta diversity

@@

### Differential abundance

@@

To estimate the proportion of variation in CH~4~ oxidation rate due to
variation in microbiome composition--i.e., microbiability (as described in the Methods; [@he2022])--we regressed divergence between the
positive selection treatment and the control against the cumulative selection
differential. The microbiability was `r myround(tidy(rh_model)[[2, 2]], 2)` &pm;
`r myround(tidy(rh_model)[[2, 3]], 2)`, though this was not significant (F~`r glance(rh_model)$df`,`r glance(rh_model)$df.residual`~ = 
`r myround(glance(rh_model)$statistic, 2)`, 
p = `r myround(glance(rh_model)$p.value, 2)`).

@@

Bray-Curtis dissimilarity of the soil microbiome varied strongly by passage and
weakly by treatment with an interaction between passage and treatment (`r figures("beta", display = "cite")`). 
Passage explained `r myround(100*tidy_permanova[1, 4])`\% of the
variation in Bray-Curtis dissimilarity (F~`r tidy_permanova[1, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[1, 5], 1)`, p = `r myround(tidy_permanova[1, 6], 3)`), treatment explained 
`r myround(100*tidy_permanova[2, 4])`\%
of the variation (F~`r tidy_permanova[2, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[2, 5], 1)`, p = `r myround(tidy_permanova[2, 6], 3)`), and the interaction
between treatment and passage explained 
`r myround(100*tidy_permanova[3, 4])`\% of the variation (F~`r tidy_permanova[3, 2]`,`r tidy_permanova[4, 2]`~ = `r myround(tidy_permanova[3, 5], 1)`, p = `r myround(tidy_permanova[3, 6], 3)`). There was 
no difference in dispersion between treatments or passages (F~3,44~ = 0.91, p = 0.45). 
Finally, CH~4~ oxidation rate was correlated with Bray-Curtis dissimilarity across both treatments in passage 5 and explained `r myround(dbrda_percent_explained[1], 1)`\% of the variation in Bray-Curtis dissimilarity (dbRDA: F~`r dbrda_aov[1, 1]`,`r dbrda_aov[2, 1]`~ = `r myround(dbrda_aov[1, 3], 2)`, p = `r myround(dbrda_aov[1, 4], 2)`)

@@

We did not identify any methanotrophic families in the overall differential abundance analysis. 
However, we wanted to look more closely at the known methanotrophs in our dataset to be sure that they did not have an effect.
To do this, we subset the ASVs in our dataset that were in families that contained methanotrophs.
Only two families were represented: *Methylacidiphilaceae* and *Beijerinckiaceae*. 
Aggregating reads at the family level, neither family was differentially abundant between the two treatments.
However, aggregated at the genus level, a group of unclassified genera in the *Beijerinckiaceae* were depleted in the selection treatment and the genus
*Rhodoblastus*, a member of the *Beijerinckiaceae*, was enriched in the
selection treatment.  While many *Beijerinckiaceae* are methanotrophs, several
taxa in this family have lost the ability to oxidize CH~4~ and it appears
that *Rhodoblastus* species are not able to grow on CH~4~, though they can grow
on methanol [@tamas2014; @dedysh2016]. Based on this analysis,
it appears that no known methanotrophs were enriched in the positive selection treatment.

@@

We used artificial ecosystem selection to estimate the contribution of variation
in microbiome composition to variation in the rate of an ecosystem function,
CH~4~ oxidation in soil, independent of environmental variation. Understanding
how and to what degree microbiome variation contributes to variation in
ecosystem function is important for many reasons. For example, successful
microbiome manipulations require that the manipulated microbiome contribute to
variation in ecosystem function independent of other drivers of ecosystem
variation (such as variation in environmental conditions). This is because the
drivers of variation in ecosystem functions can interact in complicated ways
(`r figures("concept", display = "cite")`);
for example, environmental variation can indirectly contribute by providing
conditions that select for microbial groups that in turn alter the rates of
ecosystem functions. Determining the independent contribution of microbiome
variation to ecosystem function is crucial because if microbiome composition is
driven primarily by environmental conditions, then introducing a desirable taxon
through microbiome manipulation without altering the environment will likely be
unsuccessful at shifting the targeted ecosystem function. 
The artificial selection approach is different from the comparative and
manipulative approaches used in past attempts at answering this question,
because it can control for both the direct effect of environment on function as
well as the indirect effect of environment via its impact on microbiome
assembly.

In our study, we observed an increase in CH~4~ oxidation rate in the
selection treatment relative to the control, which demonstrates that there was a
response to selection. In addition, we observed a significant correlation 
between CH~4~ oxidation rate and overall microbiome composition. Given these observations, we
conclude that variation in the microbiome contributes to variation in the
CH~4~ oxidation rate independent of the environment. This suggests that
microbiome manipulations could be an effective approach for altering the rate of
CH~4~ oxidation in this soil, and that the artificial selection approach may
be useful in determining the potential for microbiome manipulations for other
functions in other ecosystems.

Given that variation in the microbiome is associated with variation in the rate
of an ecosystem function in our system, we next wanted to estimate how
much variation in ecosystem function is associated with microbiome variation.
The relationship between microbiome variation and ecosystem function variation
is analogous to the concept of "heritability" [@lynch1998] used by quantitative
geneticists, or more precisely the concept of "microbiability" [@he2022] proposed
by microbiome scientists who study host-associated microbiomes. Although rarely
used in the study of environmental microbiomes, this concept could be
useful for understanding and manipulating microbially-mediated functions in a
variety of ecosystems.

In our experiment, the microbiability was not statistically
significant and, therefore, we were not able to accurately estimate
the percentage of variation in CH~4~ oxidation rate explained by the microbiome. 
However, we did observe a significant divergence between the 
positive selection and control lines as a result of passaging the microbiome.
This suggests that the imposed
selection was sufficient to generate
variation in soil CH~4~ oxidation rate via the microbiome. Future studies with
greater replication could more precisely estimate the microbiability.
Even though we attempted to estimate the microbiability in this study, 
it is likely that the
“environmental microbiability” will be different for other ecosystem functions
in this soil and for CH~4~ oxidation in other soils. However, our experiment
provides an example of how
this relationship can be measured

@@

# Conclusions

There is increasing interest in using artificial selection for understanding and
manipulating the microbiomes associated with plants and animals (a.k.a., “microbiome
breeding”; @mueller2022). Our study demonstrates that artificial ecosystem
selection can also be an important tool for exploring the relationship between
microbiome composition and ecosystem function in non-host systems. This
approach can provide unique information about the independent contribution of
microbiomes to ecosystem functions. For example, we demonstrated that passaging
of soil microbiomes can produce a response to selection on soil CH~4~ oxidation
rate. This demonstrates that microbiome variation can contribute to variation in 
ecosystem function independent of the environment.
Such information is crucial if we are to
successfully manipulate environmental microbiomes to alter ecosystem functions,
whether to improve crop productivity [@chaparro2012]  or
ameliorate the impacts of environmental change [@jansson2020].

# Acknowledgments {.unnumbered}

@@

# Competing interests {.unnumbered}

@@

# Data availability statement {.unnumbered}

@@

# References {.unnumbered}

@@

# Figure Captions {.unnumbered}
